---
title: ""
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: no
    toc: yes
    toc_float: yes
  pdf_document:
    number_sections: no
    toc: yes
---


# Food security of vulnerable Amazonian populations affected by large hydroelectric dams {.tabset}


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
source("HighstatLibV10.R")
library(lattice)
library(INLA)
library(ggplot2)
library("bayesplot")
library("ggplot2")
library("rstanarm")
```


## Data {.tabset}


```{r, warning=FALSE,message=FALSE}
library(readxl)
data<- read_excel("data_yes.xlsx",sheet = 1)
data$DATA=as.Date(data$DATA,format="%y-%m-%d")
data$TOTAL_RIQUEZA_PEIXES=as.numeric(data$TOTAL_RIQUEZA_PEIXES)
data$`Total_Peixes(g)`=as.numeric(data$`Total_Peixes(g)`)
head(data)
```

**Including river level rate covariable:** code for merging both data sets.


```{r message=FALSE, warning=FALSE, paged.print=TRUE}
vazao <- read_excel("Dados_Cota_BM.xlsx",sheet=5)

vazao$DATA=as.Date(vazao$DATA,format="%y-%m-%d")

names_d=unique(data$INFLUENCIA)
names_v=unique(vazao$cod)



aa=subset(vazao, vazao$cod=="Jusante")
aa$cod=rep("DOWNSTREAM_RURAL",dim(aa)[1])
vazao=rbind(vazao,aa)


ind1=which(vazao$cod=="Jusante")
vazao[ind1,3]="DOWNSTREAM_URBAN"

ind1=which(vazao$cod=="Reservatório")
vazao[ind1,3]="RESERVOIR_RURAL"


ind1=which(vazao$cod=="VoltaGrande")
vazao[ind1,3]="DEWATERED_REACH_RURAL"

ind1=which(vazao$cod=="RESERVOIR_RURAL")
aux=vazao[ind1,]
aux[,3]="RESERVOIR_URBAN"
vazao=rbind(vazao, aux)

colnames(vazao)=c("Codigo",	"Estacion",	"cod",	"DATA",	"ANO",	"MES",	"Cota(cm)")
vazao$newdate=(vazao$DATA)
data$newdate=(data$DATA)
l=dim(data)[1]

data$vazao=1:l
j=dim(data)[2]

for(i in 1:l){
  f=data[i,]$newdate
  a=data[i,]$INFLUENCIA
  b=subset(vazao,vazao$newdate==f)
  c=subset(b,b$cod==a)
  if(dim(c)[1]>0) {data[i,j]=c$`Cota(cm)`} else {data[i,j]=NA}
}
head(data)
```


## Previous steps {.tabset}

### Check TOTAL_RIQUEZA_PEIXES {.tabset}

<span style="color:blue">After replacing 0 by NA when `Total_Peixes(g)`>0 and 1 by NA when `Total_Peixes(g)`=0.</span>


- Values of  TOTAL_RIQUEZA_PEIXES if `Total_Peixes(g)`=0.
```{r}
a=subset(data,data$`Total_Peixes(g)`==0)
summary(a$TOTAL_RIQUEZA_PEIXES)
```
- Values of  TOTAL_RIQUEZA_PEIXES if `Total_Peixes(g)`>0.

```{r}
b=subset(data,data$`Total_Peixes(g)`>0)
summary(b$TOTAL_RIQUEZA_PEIXES)
table(b$TOTAL_RIQUEZA_PEIXES)
```


In previous analysis the values of TOTAL_RIQUEZA_PEIXES>=3 had been eliminated.

```{r}
dim(data)

ind=which(data$TOTAL_RIQUEZA_PEIXES==3)
ind1=which(data$TOTAL_RIQUEZA_PEIXES==4)
ind2=which(data$TOTAL_RIQUEZA_PEIXES==5)

data=data[-c(ind1,ind,ind2),]
table(data$TOTAL_RIQUEZA_PEIXES)
```

### Defining response variables (dealing with the dependency problem) {.tabset}

Presence/absence of fish consumption.

```{r}
data$pres<-ifelse(data$`Total_Peixes(g)`>0,1,0)
table(data$pres)
```

Fish intake.

```{r}
data$consumo=data$`Total_Peixes(g)`/data$NUMERO.DE.PESSSOAS
```

We compute its mean by groups. We take into account in the process the different categories of "Localidade" and "Influencia" covariables.

```{r,message=FALSE, warning=FALSE}
library(dplyr)

f=function(x){mean(x,na.rm=TRUE)}
data1=data %>%
  group_by(DATA,LOCALIDADE,INFLUENCIA) %>%
  summarise_at(vars(c(pres,vazao,consumo,TOTAL_RIQUEZA_PEIXES)), f)
head(data1)

# Number of zeros
sum(data1$pres==0)/length(data1$pres)
# Number of 1's
sum(data1$pres==1)/length(data1$pres)
```

### Presence/absence mean {.tabset}

The most appropriate model for the mean of presence/absence of fish consumption is a beta model with a logistic
link as it will ensure that fitted values are always between 0 and 1. The crucial point is that the response variable must take values between 0
and 1, excluding the boundaries. Hence it is necessary to apply a 
simple transformation 
(Smithson and Verkuilen, 2006):

$$(Y_i × (N – 1) + 0.5) / N,$$
where $N$ is the sample size.


```{r}
N <- nrow(data1)
data1$pres_bound=(data1$pres * (N - 1) + 0.5 ) / N
summary(data1$pres_bound)
boxplot(data1$pres_bound,col="gold")
```

### Exploratory of consumo response variable {.tabset}


```{r}
summary(data1$consumo)
sum(data1$consumo==0)/length(data1$consumo)
boxplot(data1$consumo,col="gold")
```

### Defining time and season covariables {.tabset}

Seasonal variable


```{r, message=FALSE}

data1$newdate=(data1$DATA)
data1$season=yday(data1$newdate)
# Interesting link: https://data.library.virginia.edu/working-with-dates-and-time-in-r-us## ing-the-lubridate-package/
summary(data1$season)
```

Continuous time variable

```{r}
data1$Time <- julian(data1$newdate, origin = as.Date("2012-07-17"))
summary(data1$Time)
```



### Exploratory covariables {.tabset}


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
Myvar=c(6,8,5,7,10,11)

corvif(data1[,Myvar[-(1:2)]])


Mypairs(data1[,Myvar[-c(1,4)]])
data0=subset(data1,data1$consumo>0)
Mypairs(data0[,Myvar[-2]])
cor.test(data1$vazao,data1$season)
par(mfcol=c(2,2))
boxplot(data1[,Myvar[3]],main="River level",col="gold")
boxplot(data1[,Myvar[4]],main="TOTAL_RIQUEZA_PEIXES",col="gold")
boxplot(data1[,Myvar[5]],main="season",col="gold")
boxplot(data1[,Myvar[6]],main="Time",col="gold")


```





```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
table(data1$INFLUENCIA)
l=dim(data1)[1]
library(ggplot2)
ggplot(data1, aes(x=INFLUENCIA, y=consumo ,fill=INFLUENCIA)) + 
  geom_boxplot()+ theme(axis.title.x=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank())

```







## Models {.tabset}


```{r,message=FALSE,warning=FALSE}
library(INLA)
```






```{r,warning=FALSE,message=FALSE, include=FALSE}
### GAMs {.tabset}


## 1. For beta model.

#Basis dimension (k) checking results. Low p-value (k-index<1) may
#indicate that k is too low, especially if edf is close to k'.
library(mgcv)
G1 <- gam(data1$pres_bound ~ s(data1$Time, k =-1)+s(data1$vazao, k =-1)+data1$INFLUENCIA, data = data1, family=betar(link="logit"))
summary(G1)
a=k.check(G1)
a
plot(G1)


```







```{r,include=FALSE}
# 2. For gamma model.
data1$consumopos <- ifelse(data1$consumo>0,data1$consumo,NA)
data1$TOTAL_RIQUEZA_PEIXES<-ifelse(data1$TOTAL_RIQUEZA_PEIXES>0,data1$TOTAL_RIQUEZA_PEIXES,NA)
G1 <- gam(data1$consumopos ~ s(data1$Time, k =-1)+s(data1$vazao, k =-1)+data1$INFLUENCIA+s(data1$TOTAL_RIQUEZA_PEIXES, k =-1), data = data1, family=Gamma(link = "log"))
summary(G1)
a=k.check(G1)
a
plot(G1)
```


### INLA models {.tabset}


```{r}
# Define the new INFLUENCIA variable
influ=(data1$INFLUENCIA)
influ[which(influ=="DEWATERED_REACH_RURAL")] <- NA
influ=as.factor(influ)

N <- nrow(data1)
Xgamma <- data.frame(Intercgamma = rep(1, N), 
                    INFLUENCIAgamma   = influ,
                    TOTAL_RIQUEZA_PEIXESgamma=data1$TOTAL_RIQUEZA_PEIXES,
                    seasongamma=data1$season,
                    Timegamma=data1$Time,
                    vazaogamma=data1$vazao,
                    grouped_Timegamma = inla.group(data1$Time,n=10),
                    grouped_seasongamma =  inla.group(data1$season,n=10),
                    grouped_vazaogamma = inla.group(data1$vazao,n=8),
                    grouped_riqueza_gamma =inla.group(data1$TOTAL_RIQUEZA_PEIXES,n=15))


Xbeta <- data.frame(Intercbeta = rep(1, N), 
                    INFLUENCIAbeta   = influ,
                    TOTAL_RIQUEZA_PEIXESbeta=data1$TOTAL_RIQUEZA_PEIXES,
                    seasonbeta=data1$season,
                    Timebeta=data1$Time,
                    vazaobeta=data1$vazao,
                    grouped_Timebeta = inla.group(data1$Time,n=10),
                    grouped_seasonbeta = inla.group(data1$season,n=10),
                    grouped_vazaobeta = inla.group(data1$vazao,n=8),
                    grouped_riqueza_beta = inla.group(data1$TOTAL_RIQUEZA_PEIXES,n=15) )
```



```{r}


Stackgamma <- inla.stack(
  tag  = "FitGamma",
  data = list(AllY = cbind(data1$consumopos, NA)),  
  A    = list(1),                      
  effects = list(                 
    list(Xgamma = Xgamma))) 


Stackbeta <- inla.stack(
  tag  = "FitBeta",
  data = list(AllY = cbind(NA, data1$pres_bound)),  
  A    = list(1),                      
  effects = list(                 
    list(Xbeta = Xbeta)))

# Combined stack
Stack4 <- inla.stack(Stackgamma,Stackbeta)
```





#### Model without vazao  {.tabset}

```{r}
f4 <- AllY~ -1+
  Intercbeta+
  f(INFLUENCIAbeta, model = "iid",hyper = list(prec =    list(initial    =log(10^{-5}), fixed =TRUE))) + 
  f(grouped_Timebeta,model = "rw2")+
 # f(grouped_vazaobeta, model = "rw2")+
  f(grouped_seasonbeta,cyclic=T, model = "rw2")+
  
  
  Intercgamma + 
  f(INFLUENCIAgamma, model = "iid",hyper = list(prec =   list(initial =     log(10^{-5}),fixed=TRUE)))+
  f(grouped_Timegamma, model = "rw2")+
 # f(grouped_vazaogamma, model = "rw2")+
  f(grouped_riqueza_gamma, model = "rw2")+
  f(grouped_seasongamma,cyclic=T, model = "rw2")
  
I4 <- inla(f4,
           family = c("gamma", "beta"),
           data = inla.stack.data(Stack4),
           control.compute = list(dic = TRUE, waic = TRUE,return.marginals.predictor=TRUE),
           control.predictor = list(
             A = inla.stack.A(Stack4),compute=TRUE),verbose = TRUE,num.threads = 1
            )
summary(I4)
vec1=c(I4$waic$waic,
I4$dic$dic)
round(I4$summary.fixed,3)
I4$summary.random$INFLUENCIAbeta
I4$summary.random$INFLUENCIAgamma
```


#### Model without INFLUENCIA  {.tabset}

```{r}
f5 <- AllY~ -1+
  Intercbeta+
  #f(INFLUENCIAbeta, model = "iid",hyper = list(prec =    list(initial    =log(10^{-5}), fixed =TRUE))) + 
  f(grouped_Timebeta,model = "rw2")+
  f(grouped_vazaobeta, model = "rw2")+
  f(grouped_seasonbeta,cyclic=T, model = "rw2")+
  
  
  Intercgamma + 
  #f(INFLUENCIAgamma, model = "iid",hyper = list(prec =   list(initial =   log(10^{-5}),fixed=TRUE)))+
  f(grouped_Timegamma, model = "rw2")+
  f(grouped_vazaogamma, model = "rw2")+
  f(grouped_riqueza_gamma, model = "rw2")+
  f(grouped_seasongamma,cyclic=T, model = "rw2")
  
I5 <- inla(f5,
           family = c("gamma", "beta"),
           data = inla.stack.data(Stack4),
           control.compute = list(dic = TRUE, waic = TRUE,return.marginals.predictor=TRUE),
           control.predictor = list(
             A = inla.stack.A(Stack4),compute=TRUE),verbose = TRUE,num.threads = 1
            )
summary(I5)
vec2=c(I5$waic$waic,
I5$dic$dic)
round(I5$summary.fixed,3)
```



#### Model without vazao and Time linear  {.tabset}

```{r}
f6 <- AllY~ -1+
  Intercbeta+
  f(INFLUENCIAbeta, model = "iid",hyper = list(prec =    list(initial    =log(10^{-5}), fixed =TRUE))) + 
  #f(grouped_Timebeta,model = "rw2")+
  Timebeta+
  #f(grouped_vazaobeta, model = "rw2")+
  f(grouped_seasonbeta,cyclic=T, model = "rw2")+
  
  
  Intercgamma + 
  f(INFLUENCIAgamma, model = "iid",hyper = list(prec =   list(initial =   log(10^{-5}),fixed=TRUE)))+
  #f(grouped_Timegamma, model = "rw2")+
  Timegamma+
  #f(grouped_vazaogamma, model = "rw2")+
  f(grouped_riqueza_gamma, model = "rw2")+
  f(grouped_seasongamma,cyclic=T, model = "rw2")
  
I6 <- inla(f6,
           family = c("gamma", "beta"),
           data = inla.stack.data(Stack4),
           control.compute = list(dic = TRUE, waic = TRUE,return.marginals.predictor=TRUE),
           control.predictor = list(
             A = inla.stack.A(Stack4),compute=TRUE),verbose = TRUE,num.threads = 1
            )
summary(I6)
vec3=c(I6$waic$waic,
I6$dic$dic)
round(I6$summary.fixed,3)
I6$summary.random$INFLUENCIAbeta
I6$summary.random$INFLUENCIAgamma
```


#### Model without vazao and Season {.tabset}

```{r}
f7 <- AllY~ -1+
  Intercbeta+
  f(INFLUENCIAbeta, model = "iid",hyper = list(prec =    list(initial    =log(10^{-5}), fixed =TRUE))) + 
  f(grouped_Timebeta,model = "rw2")+
 # f(grouped_vazaobeta, model = "rw2")+
 # f(grouped_seasonbeta,cyclic=T, model = "rw2")+
  
  
  Intercgamma + 
  f(INFLUENCIAgamma, model = "iid",hyper = list(prec =   list(initial =   log(10^{-5}),fixed=TRUE)))+
  f(grouped_Timegamma, model = "rw2")+
 # f(grouped_vazaogamma, model = "rw2")+
  f(grouped_riqueza_gamma, model = "rw2")
 # f(grouped_seasongamma,cyclic=T, model = "rw2")
  
I7 <- inla(f7,
           family = c("gamma", "beta"),
           data = inla.stack.data(Stack4),
           control.compute = list(dic = TRUE, waic = TRUE,return.marginals.predictor=TRUE),
           control.predictor = list(
             A = inla.stack.A(Stack4),compute=TRUE),verbose = TRUE,num.threads = 1
            )
summary(I7)
vec4=c(I7$waic$waic,
I7$dic$dic)
round(I7$summary.fixed,3)
I7$summary.random$INFLUENCIAbeta
I7$summary.random$INFLUENCIAgamma
```

#### Model without INFLUENCIA and season  {.tabset}

```{r}
f8 <- AllY~ -1+
  Intercbeta+
  #f(INFLUENCIAbeta, model = "iid",hyper = list(prec =    list(initial    =log(10^{-5}), fixed =TRUE))) + 
  f(grouped_Timebeta,model = "rw2")+
  f(grouped_vazaobeta, model = "rw2")+
  #f(grouped_seasonbeta,cyclic=T, model = "rw2")+
  
  
  Intercgamma + 
  #f(INFLUENCIAgamma, model = "iid",hyper = list(prec =   list(initial =   log(10^{-5}),fixed=TRUE)))+
  f(grouped_Timegamma, model = "rw2")+
  f(grouped_vazaogamma, model = "rw2")+
  f(grouped_riqueza_gamma, model = "rw2")
 # f(grouped_seasongamma,cyclic=T, model = "rw2")
  
I8 <- inla(f8,
           family = c("gamma", "beta"),
           data = inla.stack.data(Stack4),
           control.compute = list(dic = TRUE, waic = TRUE,return.marginals.predictor=TRUE),
           control.predictor = list(
             A = inla.stack.A(Stack4),compute=TRUE),verbose = TRUE,num.threads = 1
            )
summary(I8)
vec5=c(I8$waic$waic,
I8$dic$dic)
round(I8$summary.fixed,3)
```



#### Summary {.tabset}

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
measures=rbind(vec1,vec2,vec3,vec4,vec5)
colnames(measures)=c("waic","dic")
rownames(measures)=c("Without vazao","Without Influencia", "Without vazao and linear time", "without vazao and season","without influencia and season")
measures
```


Best model the model without vazao.



### Plots {.tabset}


#### Trend (beta) {.tabset}


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

library(ggplot2)
suabinm <- I4$summary.random$grouped_Timebeta$mean
suabin2 <- I4$summary.random$grouped_Timebeta$`0.025quant`
suabin9 <-I4$summary.random$grouped_Timebeta$`0.975quant`
suabinID<-I4$summary.random$grouped_Timebeta$ID
suabin<-data.frame(suabinm, suabin2,suabin9,suabinID)
m3<-ggplot(data = suabin, aes(x = suabinID, y = suabinm))+
  geom_line(aes(x = suabinID, y = suabinm), color="aquamarine3", size=0.9)+
  geom_ribbon(aes(ymin=suabin2, ymax=suabin9) ,fill="aquamarine1", alpha=0.1)+
  ggtitle(" ")+
  xlab("Time")+
  ylab("Time effect (beta) ") + theme(plot.title = element_text(hjust=0.5))+ theme_classic()
jpeg("1.jpeg", width = 1400, height = 1400, res = 300)
m3
dev.off()
```


#### Trend (gamma) {.tabset}


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

library(ggplot2)
suabinm <- I4$summary.random$grouped_Timegamma$mean
suabin2 <- I4$summary.random$grouped_Timegamma$`0.025quant`
suabin9 <-I4$summary.random$grouped_Timegamma$`0.975quant`
suabinID<-I4$summary.random$grouped_Timegamma$ID
suabin<-data.frame(suabinm, suabin2,suabin9,suabinID)
m3<-ggplot(data = suabin, aes(x = suabinID, y = suabinm))+
  geom_line(aes(x = suabinID, y = suabinm), color="darkorange", size=0.9)+
  geom_ribbon(aes(ymin=suabin2, ymax=suabin9) ,fill="darkgoldenrod1", alpha=0.1)+
  ggtitle("")+
  xlab("Time")+
  ylab("Time effect (gamma) ")+
  theme_light() + theme(plot.title = element_text(hjust=0.5))+theme_classic()
jpeg("2.jpeg", width = 1400, height = 1400, res = 300)
m3
dev.off()
```


#### Seasonal (beta) {.tabset}

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

library(ggplot2)
suabinm <- I4$summary.random$grouped_seasonbeta$mean
suabin2 <- I4$summary.random$grouped_seasonbeta$`0.025quant`
suabin9 <-I4$summary.random$grouped_seasonbeta$`0.975quant`
suabinID<-I4$summary.random$grouped_seasonbeta$ID
suabin<-data.frame(suabinm, suabin2,suabin9,suabinID)
m3<-ggplot(data = suabin, aes(x = suabinID, y = suabinm))+
  geom_line(aes(x = suabinID, y = suabinm), color="darkgreen", size=0.9)+
geom_ribbon(aes(ymin=suabin2, ymax=suabin9) ,fill="darkolivegreen3", alpha=0.1)+
  ggtitle(" ")+
  xlab("Day")+
  ylab("Season effect (beta) ")+
  theme_light() + theme(plot.title = element_text(hjust=0.5))+theme_classic()
jpeg("3.jpeg", width = 1400, height = 1400, res = 300)
m3
dev.off()
```


#### Seasonal (gamma) {.tabset}

```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

library(ggplot2)
suabinm <- I4$summary.random$grouped_seasongamma$mean
suabin2 <- I4$summary.random$grouped_seasongamma$`0.025quant`
suabin9 <-I4$summary.random$grouped_seasongamma$`0.975quant`
suabinID<-I4$summary.random$grouped_seasongamma$ID
suabin<-data.frame(suabinm, suabin2,suabin9,suabinID)
m3<-ggplot(data = suabin, aes(x = suabinID, y = suabinm))+
  geom_line(aes(x = suabinID, y = suabinm), color="violet", size=0.9)+
  geom_ribbon(aes(ymin=suabin2, ymax=suabin9) ,fill="violet", alpha=0.1)+
  ggtitle(" ")+
  xlab("Day")+
  ylab("Season effect (gamma) ")+
  theme_light() + theme(plot.title = element_text(hjust=0.5))+theme_classic()
jpeg("4.jpeg", width = 1400, height = 1400, res = 300)
m3
dev.off()
```



#### Total riqueza {.tabset}


```{r echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}

library(ggplot2)
suabinm <- I4$summary.random$grouped_riqueza_gamma$mean
suabin2 <- I4$summary.random$grouped_riqueza_gamma$`0.025quant`
suabin9 <-I4$summary.random$grouped_riqueza_gamma$`0.975quant`
suabinID<-I4$summary.random$grouped_riqueza_gamma$ID
suabin<-data.frame(suabinm, suabin2,suabin9,suabinID)
m3<-ggplot(data = suabin, aes(x = suabinID, y = suabinm))+
  geom_line(aes(x = suabinID, y = suabinm), color="gold", size=0.9)+
  geom_ribbon(aes(ymin=suabin2, ymax=suabin9) ,fill="gold", alpha=0.15)+
  ggtitle(" ")+
  xlab("Richness")+
  ylab("Richness effect (beta) ")+
  theme_light() + theme(plot.title = element_text(hjust=0.5))+theme_classic()
jpeg("5.jpeg", width = 1400, height = 1400, res = 300)
m3
dev.off()
```


## References  {.tabset}

- Smithson M, Verkuilen J (2006). A better lemon squeezer? Maximum
likelihood regression with beta-distributed dependent variables.
Psychological Methods 11: 54–71.

